Accelerated convergence of convective simulations using boundary value problems

### 2018-02-14 tex/bvp_paper.tex
rewrite of intro; needs lots of polish, but has more info/structure.

\section{Introduction}
\label{sec:intro}
Natural convection occurs in the presence of disparate timescales. Granules on the
solar surface overturn on the order of 10 minutes, whereas deep motions in the Sun at
low Mach number are constrained by the solar rotation rate of $\sim$1 month.  
These dynamical timescales are vastly shorter than the Sun's 
Kelvin-Helmholtz timescale of nearly $3 \times 10^7$ \emph{years} \cite{stix2003}. 
Modern simulations aim to model natural convection
by increasing into the high-Rayleigh-Number (Ra) regime, in which energy transportation occurs on
timescales much larger than dynamical timescales \cite{anders&brown2017}.
In this regime, simulations must pass through a great number
of convective timescales to reach a thermally converged state.
Furthermore, as Ra increases and diffusivities correspondingly decrease, 
motions become more turbulent. To properly resolve these turbulent motions, fine grid meshes
and short timesteps are required; these requirements increase the computational
cost of evolving through a single convective overturn time.
In short, achieving thermally converged, high-Ra, astrophysically interesting
simulations is an intractably expensive problem using modern numerical tools.

Prior studies of stratified convection in which a convective layer lies between stable layers
have used the knowledge of Mixing Length Theory (MLT) to adjust the initial thermodynamic structure of
the atmosphere to a state which is closer to the adiabat chosen by convection \cite{brandenburg&all2005}.
However, many modern studies of convection do not contain stable layers above and below the convection
zone, and the presence of hard boundaries and the boundary layers that they form make it difficult to
know the correct evolved adiabat \emph{a priori}.

Here we present a method for using simple boundary value problems (BVPs), 
along with information about the evolved flow fields,
to fast-forward the slow thermal evolution of convecting simulations.  
We run two sets of experiments: one in which we allow convective simulations to evolve for a
full thermal timescale before taking measurements, and another in which we employ a fast,
BVP technique which occurs on dynamical timescales. We compare these two sets of simulations to
show the validity of the BVP technique.  We use the BVP technique to run simulations
at high Ra, in the regime where running for thermal timescales becomes computationally intractable.

@@

\section{Results}
While Fig. \ref{fig:time_trace} shows us that the post-BVP and post-rundown solutions
exhibit similar system energies and fluxes, it is important to examine other aspects of
the evolved solution. The BVP mainly serves to adjust the thermodynamic state of the solution;
in \RB convection, this means that the evolved temperature profile is a good measure of whether
or not the BVP achieves the proper solution.

@@

\section{Discussion \& Conclusions}
\label{sec:results}
The method presented here is a first step towards taking meaningful measurements
of highly turbulent convection on manageable, human timescales.  As demonstrated in Figs.
1-4, this BVP method quickly converges simulations to within a few percent of the true final
state. Furthermore, the total simulation time required to use the BVPs does not change drastically from
low $S$ to high $S$, so the main prohibition on high Ra states is the difficulty of timesteps becoming
increasingly small as turbulence increases.  Additionally, post-BVP, we generally see an increase of the
timestep size by nearly a factor of two due to the less intense driving of the near-converged state that
is achieved quickly.

>>>>

\section{Introduction}
\label{sec:intro}
Natural convection occurs in the presence of disparate timescales which
prohibit numericists from studying realistic models of natural systems.  For example,
flows in the Sun's convection zone range from a Mach number (Ma) of order unity
near the solar surface to order $10^{-5}$ deep in the solar interior.
Explicit methods which are bound by the Courant-Friedrich-Lewy
(CFL) timestep limit must resolve the fastest motions (sound
waves and surface convection), resulting in timesteps which are prohibitively
small for studies of the deep, low-Ma motions. These systems are numerically
stiff. Even in the absence of the fast surface motions, the disparity between
the sound crossing time and the deep flows have made studies of low-Ma stellar
convection difficult. Traditionally, approximations such as
the anelastic approximation, in which sound waves are explicitly filtered out,
have been used to study low-Ma flows \cite{brown&all2010, featherstone&hindman2016}.
More recently, advanced numerical techniques which use implicit or mixed
implicit-explicit timestepping mechanisms have made it feasible to study
convection at low Mach numbers \cite{viallet&all2011, viallet&all2013, viallet&all2016, lecoanet&all2014,
anders&brown2017, bordwell&all2018}, and careful studies of deep convection which
would have been impossible a decade ago are now widely accessible.

While solutions to problem of disparate dynamical timescales have proven useful,
the difference between convective timescales and system relaxation timescales remains
a significant problem facing studies of convection. 
As modern simulations aim to model natural convection
by increasing into the high-Rayleigh-number (Ra) regime,
the thermal diffusion timescale on which the system relaxes grows with respect
to the dynamical timescale, which stays roughly constant across all Ra
\cite{anders&brown2017}.
Solar convection is, once again, a prime example of this phenomenon.
Dynamical timescales in the Solar convective zone are relatively short 
(10 min overturn at solar surface, one month solar
rotation rate) compared to the Kelvin-Helmholtz timescale of
$3 \cdot 10^7$ years \cite{stix2003}.  
Furthermore, as dynamical and diffusive timescales separate, 
simulations become more turbulent. More turbulent motions 
require finer grid meshes and smaller timesteps
to capture advective dynamics. Thus, the progression of simulations into the high-Ra
regime of natural convection is slowed by two simultaneous effects: timestepping
through a single convective overturn time becomes more computationally expensive
while the number of overturn times required for systems to reach thermal equilibration
grows.

The vast difference between convective and diffusive timescales has long plagued
numericists studying convection, and a plethora of approaches has been employed to
study thermally converged solutions. One popular method for accelerating the convergence
of high-Ra solutions is by ``Bootstrapping,'' or the process of using the flow
fields in a converged solution at lower Ra as initial conditions for a simulation at a higher
Ra.  This method has been used with great success \cite{johnston&doering2009, verzicco&camussi1997},
but it is not without its faults.  Bootstrapped solutions are succeptible to hysteresis
effects, in which large-scale convective structures present in the
low Ra solution imprint onto the dynamics of the new, high Ra solution. 
In moderate-Ra simulations, a commonly-used tactic is to use 
a simple model of the full convective state as initial conditions.  
For example, past studies have used a linear eigenvalue solve to set the initial
convective state \cite{hurlburt&all1984} or used an axisymmetric solution 
as initial conditions for convection in a 3D cylinder \cite{verzicco&camussi1997}. 
In certain systems, 
the approximate state of the evolved solution can be estimated, and there an
appropriate set of initial conditions can either be solved for analytically
\cite{couston&all2017} or by using knowledge of Mixing Length Theory or other convective
theories to adjust the initial profile towards the proper adiabatic state \cite{brandenburg&all2005}.

Despite the plethora of methods that have been used,
the most straightforward way to achieve a thermally converged solution
is to evolve a convective simulation through a thermal timescale. Some modern
studies do just that \cite{featherstone&hindman2016}, but it is becoming
increasingly difficult. Such evolution is
\emph{expensive}, and state-of-the-art simulations at the highest values of Ra
can only be reasonably run
for tens to hundreds of buoyancy times \cite{stevens&all2011}, much less the
thousands of buoyancy timescales required for thermal convergence.

In this work, we study a method of achieving accelerated evolution of a
convective simulation through adjusting the thermodynamic state using information 
from resolved, convective dynamics. We couple measurements of the (non-converged)
convective solution with knowledge about energy balances in the eventual solution
to self-consistently adjust the mean thermodynamic profile towards. 
While such a method has been used previously \cite{hurlburt&all1986}, 
we find no explanation in the current literature of the steps involved in employing
this method, nor any study into the accuracy of such a method.
In section \ref{sec:experiment}, we describe our convective simulations, our
numerical methods, and our method for achieving accelerated evolution. In
section \ref{sec:results}, we compare solutions reached through the accelerated evolution
method to those that have been evolved through a full diffusive timescale, and we
examine select simulations at very high Ra which have achieved accelerated evolution. Finally,
in section \ref{sec:discussion}, we discuss extensions of the methods presented
here to stratified, compressible systems.

@@

\section{Results}
\label{sec:results}
While Fig. \ref{fig:time_trace} shows us that the post-BVP and post-rundown solutions
exhibit similar system energies and fluxes, it is important to examine other aspects of
the evolved solution. The BVP mainly serves to adjust the thermodynamic state of the solution;
in \RB convection, this means that the evolved temperature profile is a good measure of whether
or not the BVP achieves the proper solution.

@@

\section{Discussion \& Conclusions}
\label{sec:discussion}
The method presented here is a first step towards taking meaningful measurements
of highly turbulent convection on manageable, human timescales.  As demonstrated in Figs.
1-4, this BVP method quickly converges simulations to within a few percent of the true final
state. Furthermore, the total simulation time required to use the BVPs does not change drastically from
low $S$ to high $S$, so the main prohibition on high Ra states is the difficulty of timesteps becoming
increasingly small as turbulence increases.  Additionally, post-BVP, we generally see an increase of the
timestep size by nearly a factor of two due to the less intense driving of the near-converged state that
is achieved quickly.
